NumLin: Linear Types for Linear Algebra

### 2018-11-15 write-up/paper/implementation.tex
Finish Performance Metrics (for Kalman, for now)

Here, evaluate the performance of the examples from the second
section.  Compare with your C implementations, and perhaps as well as
the straightforward math transcribed into (Matlab/R/Numpy?).

@@

>>>>

We think that using \lang\ has two primary benefits: safety and performance. We
discuss safety in \ref{subsec:finding_bugs}, where we describe how we used
\lang\ to find linearity and aliasing bugs in a linear algebra algorithm that
was \emph{generated} by another program.

\subsubsection{Setup}

For performance, we measured the execution of four equivalent
implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
intended, safe/copying-by-default interface), and Python (using \textsc{NumPy},
with the interpreter started and functions interpreted). We measured execution
time in micro-seconds, against an exponentially (powers of 5) increasing
scaling factor for matrix size parameters $n=5$ and $k=3$.

For large scaling factors ($n = 5^4, 5^5$), we triggered a full
garbage-collection before measuring the execution time of a single call of a
function. However, due to the limitations of the micro-benchmarking library we
used, for smaller scaling factors ($n = 5^1, 5^2, 5^3$), we measured the
execution time of \emph{multiple} calls to a function in a loop, thus including
potential garbage-collection effects.

\subsubsection{Hypothesis}

We expected the C implementation to be faster than the \lang\ one because the
latter has the additional (but relatively low) overhead of dimension checks and
crossing the OCaml/C FFI for each call to a \textsc{Cblas} routine, even though
the calls and their order are exactly the same. We expected the OCaml and
Python implementations to be slower because they allocate more temporaries (so
possibly less cache-friendly) and carry out more floating-point operations --
the \textsc{Cblas} and \lang\ implementations use ternary kernels (coalescing
steps), a Cholesky decomposition (of a symmetric matrix, which is more
efficient than the LU decomposition used for inverting a matrix in \textsc{Owl}
and \textsc{NumPy}) and \texttt{symm} (symmetric matrix multiplication, halving
the number of floating-point multiplications required).

\subsubsection{Results}

% Need to re-run timings

The results in Figure~\ref{fig:timings} are as we expected: C is the fastest,
followed by \lang, with OCaml and Python last.  Differences in timings are
quite pronounced at small matrix sizes, but are still significant at larger
ones: for $n=625$, \textsc{Cblas} took $112 \pm 35\, ms$, \lang\ took $105 \pm
25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$ and \textsc{NumPy} took $112
\pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took $10.8 \pm 0.7\, s$, \lang\
took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3 \pm 0.2\, s$ and
\textsc{NumPy} took $12.7 \pm 0.6\, s$.

Worth highlighting here is the other major advantage of using \lang\ is reduced
memory usage.  Whilst the \textsc{Owl} and \textsc{NumPy} use 11 temporary
matrices (\emph{excluding} the 2 matrices which store the results), using $n +
n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$) words of memory,
\textsc{Cblas} and \lang\ use only \emph{2} temporary m`atrices (excluding the
\emph{one} matrix which stores one of the results), using only $n^2 + nk \leq
2n^2$ words of memory.

@@

\subsubsection{Analysis}

As matrix sizes increase, assuming sufficient memory, the difference in the
number of floating-point operations ($O(n^3))$ dominates execution times.
However for small matrix sizes, since $n$ is small and the measurements were
over multiple calls to a function in a loop, the large number of temporaries
show the adverse effect of not re-using memory at even quite small matrix
sizes: creating pressure on the garbage collector.

% Other examples impl
% Sampled linear reg example
